Liveness-Based Garbage Collection for Lazy Languages

### 2016-02-03 paper/paper.tex
compressed enum and items

@@

\definecolor{Myblue}{rgb}{.2,0,1}
\definecolor{Myred}{rgb}{1,.2,.2}
\newcommand{\mycomment}[1]{{\color{Myblue}{#1}} }
\newcommand{\added}[1]{{\color{Myblue}{#1}}}
\newcommand{\todelete}[1]{{\color{Myred}{#1}}}

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are
\added{
\begin{myitemize}
\item We propose a liveness based garbage collection scheme for a lazy
  first order functional  language and prove its  correctness.  To the
  best  of our  knowledge, this  is the  first work  on liveness-based
  garbage    collection   for    lazy    languages.    All    previous
  attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
    asati14lgc,    karkare06effectiveness}   have    involved   either
  imperative  or  eager  functional  languages.   The  novely  of  our
  approach is in  handling of  liveness of  functional values  (i.e.,
  closures), an issue that arises due to laziness.

@@

\item We have implemented a garbage  collector that uses the result of
  liveness  analysis to  retain  live cells.   Our experiments  reveal
  interesting  space-time   trade-offs  in  the  engineering   of  the
  collector---for  example, updating  liveness information  carried in
  closures during  execution results in more  garbage being collected.
  Empirical results  show the effectiveness of  liveness-based garbage
  collection.
\end{myitemize}
}
\todelete{
\begin{myitemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first-order
  functional language and  prove its correctness. The 
  proof involves specifying liveness through a non-standard semantics  and then proving the analysis 
  correct with respect to the specification.   
\item The analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether   to  copy  a  cell   during  garbage  collection
  translates  to a  membership problem  for such  grammars.  In  this
  paper, we  have shown the  undecidability of this problem and 
  circumvent the  problem by over-approximating the  grammars
  with DFAs.
\item We  have implemented a garbage  collector that uses the  DFAs to
  retain live  cells.  Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  show the effectiveness of liveness-based garbage collection.
\end{myitemize}

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.   Here are  the ideas  behind the
proof.
\begin{myenumerate}
\item       We   augment     the        standard       semantics
  (Figure~\ref{fig:lang-semantics}) to model  a dynamic liveness analysis 
that records liveness values at each reference created during execution.
Any attempt to dereference a reference \mycomment{that has been declared not live by our analysis} %%with a null-liveness   
 during  a  transition  will  result  in entering  a
  special state denoted \bang. We call the semantics after
  augmentation,  \emph{minefield semantics}.
\item \label{inline1} Assuming that a  program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function calls which has the same minefield behavior.
\item  The final  step   shows  that  no program  without
  function calls can enter a \bang\ state. As a consequence no program
  (with or without function calls) can enter the \bang\ state. 
\end{myenumerate}

@@

To set up the minefield semantics, we follow these steps:
\begin{myenumerate}

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial state is  $\sigma_{all}$, and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it to the state after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{myenumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environments for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{myenumerate}
\end{myenumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

@@

In  summary, the  equations  generated during the
analysis are:
\begin{myenumerate}
\item   For  each   function  $\mathit{f}$,   equations
  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item   For  each   function  $\mathit{f}$   (including
  $\mainpgm$) an  equation  defining
  liveness for each garbage in  $e_{\mathit f}$.
\end{myenumerate}

@@

\begin{myenumerate}
\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
  are two productions
  $\nont{i}{c} \rightarrow \overline{c'} \nont{j}{\acar}\bcar$ and $\nont{i}{c}
  \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
\end{myenumerate}
The idea  behind the  productions is explained  with an
example:  Assume that  the current  sentential form  is
$\bcar\bcdr\nont{i}{\acar}\acar\acar\$$. Also  assume that  the
TM  has  a   transition  $(S_i,\acar)  \rightarrow
(S_j,\acdr, L)$.  Since the next corresponding  step in
the  derivation  has  to  be  done  without  any  prior
knowledge of whether the symbol to the left of the tape
is a \acar\  or a \acdr, two  productions are provided,
and  the  invariant  will  be maintained  only  if  the
production         $\nont{i}{\acar}         \rightarrow
\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
in  the  derivation.    This  gives  the  configuration
$\bcar\bcdr\acdr\nont{j}{\acdr}\acdr\acar\acar\$$.
Simplification   with    the   production   $\bcdr\acdr
\rightarrow               \epsilon$              yields
$\bcar\nont{j}{\acdr}\acdr\acar\acar\$$,    which   exactly
corresponds  to   the  changed  configuration   of  the
TM.  Notice  carefully that a wrong  choice breaks
the invariant  and it cannot be  recovered subsequently
by any choice of productions.

@@

Based on  the above considerations,  we restrict the  possible garbage
collection points in a function body to the following:
\begin{myenumerate}
\item We  maintain static estimates  of the memory required  to create
  the closures for each function body.  On entering a function, if the
  available memory is  less than this requirement for  the function, a
  garbage collection is triggered.
\item Since the evaluation of a  \SIF\ condition may trigger a 
  collection, after evaluating the condition   the available memory
  is  checked once  again against  a  revised estimate  of the  memory
  (based on  value of the condition)  required to execute the  rest of
  the program.  A garbage collection is triggered if enough memory is unavailable.
\end{myenumerate}

>>>>

% to save space in lists
\usepackage{enumitem}
\setlist{nolistsep}

@@

\definecolor{Myblue}{rgb}{.2,0,1}
\definecolor{Myred}{rgb}{1,.2,.2}
\newcommand{\mycomment}[1]{{\color{Myblue}{#1}} }
\newcommand{\added}[1]{{\color{Myblue}{#1}}}
\newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}

@@

In  this work  we show  that  static analysis  of heap  data can  help
garbage  collectors   in  reclaiming   more  garbage.    The  specific
contributions of this paper are
\added{
\begin{itemize}
\item We propose a liveness based garbage collection scheme for a lazy
  first order functional  language and prove its  correctness.  To the
  best  of our  knowledge, this  is the  first work  on liveness-based
  garbage    collection   for    lazy    languages.    All    previous
  attempts~\cite{shaham01heap,  ran.shaham-sas03,  shaham02estimating,
    asati14lgc,    karkare06effectiveness}   have    involved   either
  imperative  or  eager  functional  languages.   The  novely  of  our
  approach is in  handling of  liveness of  functional values  (i.e.,
  closures), an issue that arises due to laziness.

@@

\item We have implemented a garbage  collector that uses the result of
  liveness  analysis to  retain  live cells.   Our experiments  reveal
  interesting  space-time   trade-offs  in  the  engineering   of  the
  collector---for  example, updating  liveness information  carried in
  closures during  execution results in more  garbage being collected.
  Empirical results  show the effectiveness of  liveness-based garbage
  collection.
\end{itemize}
}
\todelete{
\begin{itemize}
\item  We  formulate  a  liveness  analysis for  a  lazy  first-order
  functional language and  prove its correctness. The 
  proof involves specifying liveness through a non-standard semantics  and then proving the analysis 
  correct with respect to the specification.   
\item The analysis results  in a set  of context-free  grammars along
  with  a fixed  set  of non-context-free  productions.  The  decision
  whether   to  copy  a  cell   during  garbage  collection
  translates  to a  membership problem  for such  grammars.  In  this
  paper, we  have shown the  undecidability of this problem and 
  circumvent the  problem by over-approximating the  grammars
  with DFAs.
\item We  have implemented a garbage  collector that uses the  DFAs to
  retain live  cells.  Our  experiments reveal  interesting space-time
  trade-offs  in  the  engineering  of  the  collector---for  example,
  updating liveness  information carried in closures  during execution
  results  in   more  garbage  being  collected.    Empirical  results
  show the effectiveness of liveness-based garbage collection.
\end{itemize}

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.   Here are  the ideas  behind the
proof.
\begin{enumerate}
\item       We   augment     the        standard       semantics
  (Figure~\ref{fig:lang-semantics}) to model  a dynamic liveness analysis 
that records liveness values at each reference created during execution.
Any attempt to dereference a reference \mycomment{that has been declared not live by our analysis} %%with a null-liveness   
 during  a  transition  will  result  in entering  a
  special state denoted \bang. We call the semantics after
  augmentation,  \emph{minefield semantics}.
\item \label{inline1} Assuming that a  program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function calls which has the same minefield behavior.
\item  The final  step   shows  that  no program  without
  function calls can enter a \bang\ state. As a consequence no program
  (with or without function calls) can enter the \bang\ state. 
\end{enumerate}

@@

\noindent To set up the minefield semantics, we follow these steps:
\begin{enumerate}

\item  We  extend  the  abstract  machine  state  $\rho,S,\heap,e$  to
  $\rho,S,\heap,e,\sigma$.   We call  such a  state a  \emph{minefield
  state}.  Here $\sigma$  is the ``dynamic'' demand  on the expression
  $e$, i.e., it arises from the actual sequence of function calls that
  led  to the  evaluation  of $e$. The demand for the initial state is  $\sigma_{all}$, and each expression $e$ before a $\rightsquigarrow$ step transforms the demand in accordance with the liveness rules of Section~\ref{sec:liveness-analysis} and passes it to the state after the step.  The  information in  continuation
  frames  on the  stack $S$  are also  similarly augmented  with their
  demands.  Thus a  stacked entry now takes the form  $(\rho, \ell, e,
  \sigma)$.
\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to certain references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the objects
  pointed to by these references  during an actual garbage collection.
  This is done in the following way:
  \begin{enumerate} 
  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness environments for the expressions $e$ in each of
    the  continuation frames $(\rho,  \ell, e,  \sigma)$ in
    $S$, giving a stack of  liveness environments that we shall denote
    $\mathsf{SL}$.
  \item \emph{Garbage collection of objects pointed by root sets:} For
    the environment $\rho$ in the current state and for each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each  stack entry $(\rho,\ell,e,\sigma)$
    in $S$ 
    with  $\Lv'$   as  the   corresponding  liveness   environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.
  \item \emph{Garbage  collection of  objects in  the heap:}  For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  garbage collection
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

@@

In  summary, the  equations  generated during the
analysis are:
\begin{enumerate}
\item   For  each   function  $\mathit{f}$,   equations
  defining \Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item For each function $\mathit{f}$, an equation defining the summary
  demand $\sigma_{\mathit f}$ on $e_f$.
\item   For  each   function  $\mathit{f}$   (including
  $\mainpgm$) an  equation  defining
  liveness for each garbage in  $e_{\mathit f}$.
\end{enumerate}

@@

\begin{enumerate}
\item For each transition  $(S_i, c) \rightarrow (S_j,c',L)$, there are
  two  productions $\nont{i}{c}  \rightarrow  \acar \nont{j}{\acar}c'$
  and $\nont{i}{c} \rightarrow \acdr \nont{j}{\acdr}c'$.
\item For each  transition $(S_i, c) \rightarrow (S_j,c',R)$, there
  are two productions
  $\nont{i}{c} \rightarrow \overline{c'} \nont{j}{\acar}\bcar$ and $\nont{i}{c}
  \rightarrow \overline{c'} \nont{j}{\acdr}\bcdr$. 
\end{enumerate}
The idea  behind the  productions is explained  with an
example:  Assume that  the current  sentential form  is
$\bcar\bcdr\nont{i}{\acar}\acar\acar\$$. Also  assume that  the
TM  has  a   transition  $(S_i,\acar)  \rightarrow
(S_j,\acdr, L)$.  Since the next corresponding  step in
the  derivation  has  to  be  done  without  any  prior
knowledge of whether the symbol to the left of the tape
is a \acar\  or a \acdr, two  productions are provided,
and  the  invariant  will  be maintained  only  if  the
production         $\nont{i}{\acar}         \rightarrow
\acdr\nont{j}{\acdr}\acdr$ is chosen  for the next step
in  the  derivation.    This  gives  the  configuration
$\bcar\bcdr\acdr\nont{j}{\acdr}\acdr\acar\acar\$$.
Simplification   with    the   production   $\bcdr\acdr
\rightarrow               \epsilon$              yields
$\bcar\nont{j}{\acdr}\acdr\acar\acar\$$,    which   exactly
corresponds  to   the  changed  configuration   of  the
TM.  Notice  carefully that a wrong  choice breaks
the invariant  and it cannot be  recovered subsequently
by any choice of productions.

@@

Based on  the above considerations,  we restrict the  possible garbage
collection points in a function body to the following:
\begin{enumerate}
\item We  maintain static estimates  of the memory required  to create
  the closures for each function body.  On entering a function, if the
  available memory is  less than this requirement for  the function, a
  garbage collection is triggered.
\item Since the evaluation of a  \SIF\ condition may trigger a 
  collection, after evaluating the condition   the available memory
  is  checked once  again against  a  revised estimate  of the  memory
  (based on  value of the condition)  required to execute the  rest of
  the program.  A garbage collection is triggered if enough memory is unavailable.
\end{enumerate}
